---
title: "Data Manipulation Exercises"
subtitle: "ODA R Learning Series"
author: "Matthew Kumar"
date: "Dec 4, 2020"
output:
  rmdformats::downcute:
    self_contained: true

---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(klippy)
library(tidyverse)

## Global options

opts_knit$set(width=75)
```
```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'), color = 'darkred')
```

# Prerequisites

## Reference

You may access the presentation on Data Manipulation [here](https://r4oda.netlify.app/). 

The R-Studio dplyr cheat sheet can be viewed [here](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf). 


# Exercises

## 1. Pipe Operator

We briefly discussed the pipe operator, `%>%`,  in the presentation. 

1. What are some of the benefits of using it in your day to day work? 

<details>
<summary>Click for Answer</summary>
- Write code in a series of natural, ordered steps
- Easy to digest and reason about others code
- Neat and compact (opinion)
</details> 

2. Can you describe a scenario where it might *not* be immediately useful?

<details>
  <summary>Click for Answer</summary>
 - If you're workflow is non-linear (i.e. multi directional). 
</details> 

If you're interested in learning about different types of pipes, have a look [here.](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html)

<br>
<br>

## 2. Data Manipulation

## 2.1 Dataset 1

To complete the following exercises, you will need to copy and paste the code below into your R Studio session. This should create a data frame named `adsl` in your environment. Please verify this before moving on.

```{r, eval=F}
set.seed(8083)

adsl <- data.frame(SUBJIDN  = 1:200,
                   SEX      = sample(c('M','F', NA), 200, replace = T, prob = c(.8,.15,.05)),
                   AGE      = runif(200, min=18, max=65) %>% round(),
                   TRTP     = sample(c('DRUG','PLACEBO'), 200, replace = T),
                   SAFFL    = sample(c('Y','N', NA), 200, replace = T, prob = c(.9,.07,.03)),
                   COUNTRY  = sample(c('USA','INDIA','EU','CHINA'), 200, replace = T),
                   stringsAsFactors = F)
                   
```

## 2.2 Dataset 1 Exercises

1. Examine the structure of the code used to create `adsl`. In R, it's useful to be able to simulate data on the fly in order to test a new package or function. Type `?runif` and `?sample` at the prompt to learn about these function arguments. Can you describe, in words, how AGE and SAFFL are simulated?
<br>
<br>
2. Write a `select` query that chooses all columns in `adsl` that start with S **or** end with Y.
<br>
<br>
3. Use `adsl` to create a new data frame that consists of **numeric** variables only.
<br>
<br>
4. The `count()` function in dplyr is useful for creating frequencies. Type `?count` at the prompt to read about it. Apply it to `adsl` to answer to the following:

- Compute individual frequencies for the COUNTRY and SEX variables.

- Compute a cross-tabulation for COUNTRY and SEX while removing any rows where COUNTRY or SEX are `NA`. 
<br>
<br>
5. Compute the frequency of SAFFL.
- Request that frequencies are sorted from highest to lowest.
<br>
- How might you add the percentages with your frequencies for SAFFL? Ensure your final answer is multiplied by 100.
<br> 
<br>

## 2.3 Dataset 2 

To complete the following exercises, you will need to copy and paste the code below into your R Studio session. This should create a data frame named `visits` in your environment. Please verify this before moving on.

```{r, eval=F}
set.seed(8083)

visits <- data.frame(SUBJIDN   = sample(1:10, 30, replace = T),
                     VISITDT   = Sys.Date() + runif(30, min=-365, max=0))

```

## 2. 4 Dataset 2 Exercises

1. What is the maximum number of visits across all subjects?
<br>
<br>
2. What is the earliest visit date across all subjects?
<br>
<br>
3. What is the latest visit date across all subjects?
<br>
<br>
4. Create a data frame that keeps the earliest visit *per subject* Do this two ways:
<br>
- Use a combination of `arrange` and `distinct`
<br>
- Use the `slice` function. Type `?slice` to first learn about it.
<br>
5. Create a new column that represents the number of days between `2021-01-01` and the earliest visit date *per subject*.
- **HINT**: You can freely reference 2021-01-01 as a date constant like so: `as.Date('2021-01-01')`. 

## 3. Data Tidying

## 3.1 Dataset 3

To complete the following exercises, you will need to copy and paste the code below into your R Studio session. This should create a data frame named `adae` in your environment. Please verify this before moving on.

```{r, eval=T}
set.seed(8083)

adae <- data.frame(SUBJIDN = 1:5,
                   AEACN_SUMMARY = c('DOSE MODIFIED;;DOSE NOT CHANGED;',
                                     "DOSE REDUCED;DOSE REDUCED;DRUG WITHDRAWN;",
                                     "NOT APPLICABLE;DOSE REDUCED;;",
                                     ";;;",
                                     "DRUG WITHDRAWN;;;"),
                   stringsAsFactors = FALSE)
```

## 3.2 Dataset 3 Exercises
1. AEACN_SUMMARY is a concatenated version of the original variables AEACN1, AEACN2, AEACN3. Re-create these variables using functions from the `tidyr` package. Your result data frame should have a total of 4 columns.
<br>
<details>
  <summary>Solution</summary>
```{r, eval=FALSE}
library(tidyverse)

adae_q1 <- adae %>%
             separate(AEACN_SUMMARY,
                      into = c('AEACN1','AEACN2','AEACN3'),
                      sep = ';')
```

```{r, echo=FALSE, eval=TRUE}
adae_q1 <- adae %>%
             separate(AEACN_SUMMARY,
                      into = c('AEACN1','AEACN2','AEACN3'),
                      sep = ';')
```
</details> 
<br>
2. Use `paste()` and the result of Q1 to create a new concatenated column named `AEACN_SUMMARY2`. Use `-` as a delimiter instead of `;`.
<br>
<br>
3. Filter the data frame you created in Q1 to keep subjects who eventually had the `DRUG WITHDRAWN` 
<br>
<br>
4. Reshape the data frame you created in Q1 into long format.

